---
title: "Calibrating Hedonic Pricing Model for Private Highrise Property with GWR Method"
format: html
---

# **Calibrating Hedonic Pricing Model for Private Highrise Property with GWR Method**

## **Overview**

-   A spatial technique for building [hedonic pricing](https://www.investopedia.com/terms/h/hedonicpricing.asp) models that capture **local variations** in relationships between variables.

-   Unlike **OLS**, GWR considers **spatial non-stationarity**, showing how factors affect prices differently across locations.

-   **Application:** Condominium resale prices (2015) using **structural** and **locational** variables.

## **The Data**

Two data sets will be used in this model building exercise, they are:

-   URA Master Plan subzone boundary in shapefile format (i.e. *MP14_SUBZONE_WEB_PL*)

-   condo_resale_2015 in csv format (i.e. *condo_resale_2015.csv*)

**Installing and Loading the R packages**

::: panel-tabset
# Packages Used

-   R package for building OLS and performing diagnostics tests

    -   **olsrr**

-   R package for calibrating geographical weighted family of models

    -   **GWmodel**

-   R package for multivariate data visualisation and analysis

    -   **corrplot**

-   Spatial data handling

    -   **sf**

-   Attribute data handling

    -   **tidyverse**, especially **readr**, **ggplot2** and **dplyr**

-   Choropleth mapping

    -   **tmap**

## Code Chunk

The code chunk below uses *p_load()* of pacman package to check if the packages are installed in the computer. If they are, then they will be launched into R.

```{r}
pacman::p_load(olsrr, corrplot, ggpubr, 
               sf, sfdep, GWmodel, tmap, 
               tidyverse, gtsummary,
               performance, RColorBrewer)
```
:::

::: callout-note
The **GWmodel** package offers a range of spatial analysis tools that focus on localized data. These include geographically weighted (GW) summary statistics, principal component analysis, discriminant analysis, and different types of GW regression. Some of these methods also come in robust versions that can handle outliers.

Typically, the results or parameters produced by GWmodel are visualized as maps, making it a powerful tool for exploring spatial patterns. This exploratory step often helps guide and refine more detailed or advanced statistical analyses that follow.
:::

# **Geospatial Data Wrangling**

## **Importing geospatial data**

The geospatial data used in this hands-on exercise is called MP14_SUBZONE_WEB_PL. It is in ESRI shapefile format. The shapefile consists of URA Master Plan 2014’s planning subzone boundaries. Polygon features are used to represent these geographic boundaries. The GIS data is in svy21 projected coordinates systems.

The code chunk below is used to import *MP_SUBZONE_WEB_PL* shapefile by using *st_read()* of **sf** packages.

```{r}
mpsz = st_read(dsn = "data/geospatial", 
               layer = "MP14_SUBZONE_WEB_PL")
```

The report above shows that the R object used to contain the imported MP14_SUBZONE_WEB_PL shapefile is called *mpsz* and it is a simple feature object. The geometry type is *multipolygon*. it is also important to note that mpsz simple feature object does not have EPSG information.

## **Updating CRS information**

We will update mpsz with the correct ESPG code using the following code chunk.

```{r}
mpsz_svy21 <- st_transform(mpsz, 3414)
```

After transforming the projection metadata, you can verify the projection of the newly transformed *mpsz_svy21* by using *st_crs()* of **sf** package.

The code chunk below will be used to verify the newly transformed *mpsz_svy21*.

```{r}
st_crs(mpsz_svy21)
```

This step is showing how to find the spatial extent (or geographic boundary) of the dataset mpsz_svy21 using the *st_bbox()* function from **sf**.

*st_bbox()* extracts the minimum and maximum coordinates (in both x and y directions) that define the rectangular area covering the entire dataset.

The output will provide 4 values:

-   **xmin** – minimum x-coordinate (westmost point)

-   **ymin** – minimum y-coordinate (southmost point)

-   **xmax** – maximum x-coordinate (eastmost point)

-   **ymax** – maximum y-coordinate (northmost point)

```{r}
st_bbox(mpsz_svy21)
```

# **Aspatial Data Wrangling**

## **Importing the aspatial data**

The *condo_resale_2015* is in csv file format. The codes chunk below uses *read_csv()* function of **readr** package to import *condo_resale_2015* into R as a tibble data frame called *condo_resale*.

```{r}
condo_resale = read_csv(
  "data/aspatial/Condo_resale_2015.csv")
```

We will use *glimpse()* to display the structure so that we may examine the data.

```{r}
glimpse(condo_resale)
```

```{r}
head(condo_resale$LONGITUDE)
```

```{r}
head(condo_resale$LATITUDE)
```

The print above reveals that the values of *LONGITITUDE* and *LATITUDE* fields are in decimal degree.

```{r}
summary(condo_resale)
```

## **Converting aspatial data frame into a sf object**

The *condo_resale* tibble data frame is aspatial. We will convert it to a **sf** object. The code chunk below converts condo_resale data frame into a simple feature data frame by using *st_as_sf()* of **sf**.

```{r}
condo_resale.sf <- st_as_sf(condo_resale,
                            coords = c("LONGITUDE", "LATITUDE"),
                            crs=4326) %>%
  st_transform(crs=3414)
```

```{r}
head(condo_resale.sf)
```

The outputs now show that it is in point feature data frame.

# **Exploratory Data Analysis (EDA)**

## **EDA using statistical graphics**

We can plot the distribution of *SELLING_PRICE* by using appropriate Exploratory Data Analysis (EDA) as shown in the code chunk below.

```{r}
ggplot(data=condo_resale.sf, 
       aes(x=`SELLING_PRICE`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")
```

The figure above reveals a right skewed distribution. This means that more condominium units were transacted at relative lower prices.

Statistically, the skewed dsitribution can be normalised by using log transformation. The code chunk below is used to derive a new variable called *LOG_SELLING_PRICE* by using a log transformation on the variable *SELLING_PRICE*. It is performed using *mutate()* of **dplyr** package.

```{r}
condo_resale.sf <- condo_resale.sf %>%
  mutate(`LOG_SELLING_PRICE` = log(SELLING_PRICE))
```

Now, you can plot the *LOG_SELLING_PRICE* using the code chunk below.

```{r}
ggplot(data=condo_resale.sf, 
       aes(x=`LOG_SELLING_PRICE`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")
```

Notice that the distribution is relatively less skewed after the transformation.

## **Multiple Histogram Plots distribution of variables**

In this section, we will learn how to draw a small multiple histograms (also known as trellis plot) by using *ggarrange()* of **ggpubr** package.

The code chunk below is used to create 12 histograms. Then *ggarrange()* is used to organised these histogram into a 3 columns by 4 rows small multiple plot.

```{r}
AREA_SQM <- ggplot(data=condo_resale.sf, 
                   aes(x= `AREA_SQM`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

AGE <- ggplot(data=condo_resale.sf, 
              aes(x= `AGE`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

PROX_CBD <- ggplot(data=condo_resale.sf, 
                   aes(x= `PROX_CBD`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

PROX_CHILDCARE <- ggplot(data=condo_resale.sf, 
                         aes(x= `PROX_CHILDCARE`)) + 
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

PROX_ELDERLYCARE <- ggplot(data=condo_resale.sf, 
                           aes(x= `PROX_ELDERLYCARE`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

PROX_URA_GROWTH_AREA <- ggplot(data=condo_resale.sf, 
                               aes(x= `PROX_URA_GROWTH_AREA`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

PROX_HAWKER_MARKET <- ggplot(data=condo_resale.sf, 
                             aes(x= `PROX_HAWKER_MARKET`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

PROX_KINDERGARTEN <- ggplot(data=condo_resale.sf, 
                            aes(x= `PROX_KINDERGARTEN`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

PROX_MRT <- ggplot(data=condo_resale.sf, 
                   aes(x= `PROX_MRT`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

PROX_PARK <- ggplot(data=condo_resale.sf, 
                    aes(x= `PROX_PARK`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

PROX_PRIMARY_SCH <- ggplot(data=condo_resale.sf, 
                           aes(x= `PROX_PRIMARY_SCH`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

PROX_TOP_PRIMARY_SCH <- ggplot(data=condo_resale.sf, 
                               aes(x= `PROX_TOP_PRIMARY_SCH`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

ggarrange(AREA_SQM, AGE, PROX_CBD, PROX_CHILDCARE, 
          PROX_ELDERLYCARE, PROX_URA_GROWTH_AREA,
          PROX_HAWKER_MARKET, PROX_KINDERGARTEN, PROX_MRT,
          PROX_PARK, PROX_PRIMARY_SCH, PROX_TOP_PRIMARY_SCH,  
          ncol = 3, nrow = 4)
```

## **Drawing Statistical Point Map**

Lastly, we want to reveal the geospatial distribution condominium resale prices in Singapore. The map will be prepared by using **tmap** package.

First, we will turn on the interactive mode of tmap by using the code chunk below.

```{r}
tmap_mode("plot")
```

Next, the code chunks below is used to create an interactive point symbol map.

```{r}
tm_shape(mpsz_svy21)+
  tm_polygons() +
tm_shape(condo_resale.sf) +  
  tm_dots(fill = "SELLING_PRICE",
          fill_alpha = 0.6,
          size = 0.5,
          fill.scale = tm_scale_intervals(
            style="quantile")) +
  tm_view(set_zoom_limits = c(11,14))
```

Notice that *tm_dots()* is used instead of *tm_bubbles()*.

set_zoom_limits argument of *tm_view()* sets the minimum and maximum zoom level to 11 and 14 respectively.

Before moving on to the next section, the code below will be used to turn R display into plot mode.

```{r}
tmap_mode("plot")
```

# **Hedonic Pricing Modelling in R**

In this section, you will learn how to building hedonic pricing models for condominium resale units using *lm()* of R base.

## **Simple Linear Regression Method**

First, we will build a simple linear regression model by using *SELLING_PRICE* as the dependent variable and *AREA_SQM* as the independent variable.

```{r}
condo.slr <- lm(formula=SELLING_PRICE ~ AREA_SQM, 
                data = condo_resale.sf)
```

*lm()* returns an object of class “lm” or for multiple responses of class c(“mlm”, “lm”).

The functions *summary()* and *anova()* can be used to obtain and print a summary and analysis of variance table of the results. The generic accessor functions coefficients, effects, fitted values and residuals extract various useful features of the value returned by lm.

```{r}
summary(condo.slr)
```

-   The regression equation is y = -258121.1 + 14719x₁, meaning that SELLING_PRICE increases by about 14,719 for each unit increase in AREA_SQM.

-   R² = 0.4518, indicating the model explains about 45% of the variance in resale prices.

-   The p-value (\< 0.0001) shows the model is statistically significant, so we reject the null hypothesis that the mean alone predicts prices well.

-   Both Intercept (B₀) and AREA_SQM (B₁) have p-values \< 0.001, confirming they are significant predictors.

-   The best-fit line can be visualized in a scatterplot using *geom_smooth(method = "lm")* in **ggplot2**.

```{r}
ggplot(data=condo_resale.sf,  
       aes(x=`AREA_SQM`, y=`SELLING_PRICE`)) +
  geom_point() +
  geom_smooth(method = lm)
```

Figure above reveals that there are a few statistical outliers with relatively high selling prices.

## **Multiple Linear Regression Method**

### Visualising the relationships of the independent variables

-   Before building a multiple regression model, it is essential to confirm that independent variables are not highly correlated.

-   High correlation among predictors can distort model accuracy, a problem known as multicollinearity.

-   A correlation matrix helps visualize relationships between variables.

-   In this exercise, the **corrplot** package (and *pairs()* in R) is used to plot a scatterplot matrix showing correlations among independent variables in condo_resale.

```{r}
corrplot(cor(condo_resale[, 5:23]), 
         diag = FALSE, 
         order = "AOE",
         tl.pos = "td", 
         tl.cex = 0.5, 
         method = "number", 
         type = "upper")
```

Matrix reorder is very important for mining the hidden structure and patter in the matrix. There are four methods in corrplot (parameter order), named “AOE”, “FPC”, “hclust”, “alphabet”. In the code chunk above, AOE order is used.

From the scatterplot matrix, it is clear that *Freehold* is highly correlated to *LEASE_99YEAR*. In view of this, it is wiser to only include either one of them in the subsequent model building. As a result, *LEASE_99YEAR* is excluded in the subsequent model building.

## **Building a hedonic pricing model using multiple linear regression method**

The code chunk below using *lm()* to calibrate the multiple linear regression model.

```{r}
condo.mlr <- lm(
  formula = SELLING_PRICE ~ AREA_SQM + AGE  + 
    PROX_CBD + PROX_CHILDCARE + PROX_ELDERLYCARE +
    PROX_URA_GROWTH_AREA + PROX_HAWKER_MARKET   +
    PROX_KINDERGARTEN   + PROX_MRT  + PROX_PARK +
    PROX_PRIMARY_SCH + PROX_TOP_PRIMARY_SCH +
    PROX_SHOPPING_MALL  + PROX_SUPERMARKET + 
    PROX_BUS_STOP   + NO_Of_UNITS + 
    FAMILY_FRIENDLY + FREEHOLD, 
  data=condo_resale.sf)
summary(condo.mlr)
```

::: callout-note
The code chunk above consists of two parts:

1.  *lm()* of Base R is used to calibrate a multiple linear regression model. The model output is stored in an lm object called *condo.mlr*.

2.  *summary()* is used to print the model output.
:::

## **Revising the model**

With reference to the report above, it is clear that not all the independent variables are statistically significant. We will revised the model by removing those variables which are not statistically significant.

Now, we are ready to calibrate the revised model by using the code chunk below.

```{r}
condo.mlr1 <- lm(
  formula = SELLING_PRICE ~ AREA_SQM + AGE + 
    PROX_CBD + PROX_CHILDCARE + PROX_MRT +
    PROX_ELDERLYCARE    + PROX_URA_GROWTH_AREA +
    PROX_PARK   + PROX_PRIMARY_SCH + 
    PROX_SHOPPING_MALL  + PROX_BUS_STOP + 
    NO_Of_UNITS + FAMILY_FRIENDLY + FREEHOLD,
  data=condo_resale.sf)
summary(condo.mlr1)
```

The output above reveals that all explanatory variables are statistically significant at 95% confident level.

## **Preparing Publication Quality Table: gtsummary method**

The **gtsummary** package provides an elegant and flexible way to create publication-ready summary tables in R.

In the code chunk below *tbl_regression()* is used to create a well formatted regression report.

```{r}
tbl_regression(condo.mlr1, intercept = TRUE)
```

With **gtsummary** package, model statistics can be included in the report by either appending them to the report table by using *add_glance_table()* or adding as a table source note by using *add_glance_source_note()* as shown in the code chunk below.

```{r}
tbl_regression(condo.mlr1, 
               intercept = TRUE) %>% 
  add_glance_source_note(
    label = list(sigma ~ "\U03C3"),
    include = c(r.squared, adj.r.squared, 
                AIC, statistic,
                p.value, sigma))
```

## **Regression Diagnostics**

Regression diagnostics are methods used to check whether a regression model meets its assumptions and how well it fits the data. They help identify problems such as non-linear patterns, errors that are not normally distributed, unequal variance or data points that strongly affect the results. These checks ensure that the model’s findings are accurate and reliable.

The **olsrr** R package provides tools for performing OLS regression diagnostics:

-   comprehensive regression output

-   residual diagnostics

-   measures of influence

-   heteroskedasticity tests

-   collinearity diagnostics

-   model fit assessment

-   variable contribution assessment

-   variable selection procedures

### Multicollinearity test

Multicollinearity happens when two or more independent variables in a model are closely related. When this occurs, a change in one variable is linked to a change in another, making it hard for the model to separate their individual effects on the outcome.

Testing for multicollinearity is important in multiple linear regression because it affects how reliable and clear the model’s results are. When variables are highly correlated, the estimated coefficients can become unstable and difficult to interpret. This can cause misleading results, such as making a variable seem unimportant when it actually has a real effect.

In the code chunk below, the *check_collinearity()* of **performance** package is used to test if there are sign of multicollinearity.

```{r}
check_collinearity(condo.mlr1)
```

```{r}
mlr.vif <- check_collinearity(condo.mlr1)
plot(mlr.vif)
```

Since the VIF of the independent variables are less than 10. We can safely conclude that there are no sign of multicollinearity among the independent variables.

### Test for Non-Linearity

In multiple linear regression, it is important for us to test the assumption that linearity and additivity of the relationship between dependent and independent variables.

In the code chunk below, the *ols_plot_resid_fit()* of **olsrr** package is used to perform linearity assumption test.

```{r}
check_model(condo.mlr1,
            check = "linearity")
```

The figure above reveals that most of the data points are scattered around the 0 line, hence we can safely conclude that the relationships between the dependent variable and independent variables are linear.

### Test for Normality Assumption

The normality assumption test for multiple linear regression checks if the model’s residuals (the differences between observed and predicted values) are normally distributed. This is crucial for accurate hypothesis testing and confidence intervals. To test this, you can use visual methods like histograms and Q-Q plots of the residuals, or conduct statistical tests like Shapiro-Wilk test and Kolmogorov-Smirnov test.

In the code chunk below, *check_normality()* of performance package is used to perform normality assumption test on condo.mlr1 model.

```{r}
check_normality(condo.mlr1)
```

The print above reveals that the p-value of the normality assumption test is less than alpha value of 0.05. Hence we reject the normality assumption at 95% confident level.

-   ::: callout-note
    -   *check_normality()* calls *stats::shapiro.test* and checks the standardized residuals (or studentized residuals for mixed models) for normal distribution.

    -   Note that this formal test almost always yields significant results for the distribution of residuals and visual inspection (e.g. Q-Q plots) are preferable.
    :::

Instead of showing the test statistic, *plot()* of **see** package can be used to plot the output of *check_normality()* for visual inspection as shown below.

```{r}
plot(check_normality(condo.mlr1), 
     type = "qq")
```

Q-Q plot above below shows that majority of the data points are felt along the zero line.

Another way to check for normality assumption visual is by using *check_model()* of **performance** package as shown in the code chunk below.

```{r}
check_model(condo.mlr1,
            check= "normality")
```

The figure reveals that the residual of the multiple linear regression model

(i.e. condo.mlr1) resembles normal distribution.

### Testing for Spatial Autocorrelation

The hedonic model we try to build are using geographically referenced data. Hence, it is crucial to check for spatial autocorrelation because its presence can produce unreliable and misleading results. Traditional regression models, such as ordinary least squares (OLS), assume that observations are independent of one another. However, spatial data often violates this assumption.

Spatial autocorrelation is the correlation of a variable with itself across different spatial locations. Positive spatial autocorrelation means nearby features tend to be more similar, while negative autocorrelation means they tend to be more dissimilar. This phenomenon is based on the first law of geography: “Everything is related to everything else, but nearby things are more related than distant things”.

Ignoring spatial autocorrelation in a regression model can lead to serious statistical issues:

-   Biased and inefficient coefficient estimates: If autocorrelation is present, standard errors of the model coefficients can be wrong, leading to unreliable hypothesis tests (p-values). The model might appear more significant than it is.

-   Misleading significance tests: Standard regression models cannot distinguish between true explanatory power and the influence of spatial patterns, resulting in inaccurate p-values.

-   Model misspecification: Significant spatial autocorrelation in the regression residuals often signals that important explanatory variables are missing from the model. The spatial patterning of the residuals (over- and under-predictions) can provide clues about what these missing variables might be.

-   Inflated Type I error rates: Researchers might incorrectly reject a true null.

To test for spatial autocorrelation, We can run a Moran’s I test on the model’s residuals. Significant spatial autocorrelation in the residuals means the model is not capturing the full spatial story.

In order to perform spatial autocorrelation test, we need to export the residual of the hedonic pricing model and save it as a data frame first.

```{r}
mlr.output <- as.data.frame(condo.mlr1$residuals)
```

Next, we will join the newly created data frame with *condo_resale.sf* object.

```{r}
condo_resale.sf <- cbind(condo_resale.sf, 
                        condo.mlr1$residuals) %>%
rename(`MLR_RES` = `condo.mlr1.residuals`)
```

Next, we will use **tmap** package to display the distribution of the residuals on an interactive map.

The code churn below will turn on the interactive mode of tmap.

```{r}
tmap_mode("plot")
```

The code chunks below is used to create an interactive point symbol map.

```{r}
tm_shape(mpsz_svy21) +
  tm_polygons(fill_alpha = 0.4) +
tm_shape(condo_resale.sf) +  
  tm_dots(
    fill = "MLR_RES",
    size = 0.7,
    col = "black",
    fill.scale = tm_scale(
      n = 10,
      values = rev(
        brewer.pal(11, "RdBu")),
      style = "quantile",
      midpoint = NA),
    fill.legend = tm_legend(title = "Residuals")
  ) + 
  tm_title("LM Residuals (Quantile Classification)") +
  tm_layout(legend.outside = TRUE) +
  tm_view(set_zoom_limits = c(11,14))
```

Remember to switch back to “plot” mode before continue.

```{r}
tmap_mode("plot")
```

The figure above reveal that there is sign of spatial autocorrelation.

To proof that our observation is indeed true, Global Moran’s I test will be performed

First, we will compute the distance-based weight matrix by using *st_dist_band()* function of **sfdep**.

```{r}
condo_resale.sf <- condo_resale.sf %>% 
  mutate(
    nb = st_dist_band(
      st_geometry(geometry), 
                 upper = 1500),
    wt = st_weights(
      nb, style = "W"),
    .before = 1) 
```

Next, *global_moran_perm()* of **sfdep** package will be used to perform Moran’s I test for residual spatial autocorrelation

```{r}
set.seed(1234)
global_moran_perm(
  condo_resale.sf$MLR_RES, 
  nb = condo_resale.sf$nb, 
  wt = condo_resale.sf$wt,
  alternative = "two.sided", 
  nsim = 499)
```

The Global Moran’s I test for residual spatial autocorrelation shows that it’s p-value is less than 0.00000000000000022 which is less than the alpha value of 0.05. Hence, we will reject the null hypothesis that the residuals are randomly distributed.

Since the Observed Global Moran I = 0.14389 which is greater than 0, we can infer than the residuals resemble cluster distribution.

# **Building Hedonic Pricing Models using GWmodel**

In this section, you are going to learn how to modelling hedonic pricing using both the fixed and adaptive bandwidth schemes

## **Building Fixed Bandwidth GWR Model**

### Computing fixed bandwith

In the code chunk below *bw.gwr()* of **GWModel** is used to determine the optimal fixed bandwidth to use in the model. Notice that the argument *adaptive* is set to FALSE indicates that we are interested to compute the fixed bandwidth.

There are two possible approaches can be used to determine the stopping rule, they are CV cross-validation approach and AIC corrected (AICc) approach. We define the stopping rule using *approach* agreement.

```{r}
bw.fixed <- bw.gwr(
  formula = SELLING_PRICE ~ AREA_SQM + AGE  + 
    PROX_CBD + PROX_CHILDCARE + PROX_ELDERLYCARE    +
    PROX_URA_GROWTH_AREA + PROX_MRT + PROX_PARK + 
    PROX_PRIMARY_SCH + PROX_SHOPPING_MALL   + 
    PROX_BUS_STOP   + NO_Of_UNITS + FAMILY_FRIENDLY + 
    FREEHOLD, 
  data=condo_resale.sf, 
  approach="CV", 
  kernel="gaussian", 
  adaptive=FALSE, 
  longlat=FALSE)
```

The result shows that the recommended bandwidth is 971.3405 metres.

::: callout-note
The bandwidth value is expressed in metres because the spatial dataset adopts Singapore’s standard coordinate reference system, SVY21 (EPSG:3414), which measures all spatial distances and areas in metres. When the Geographically Weighted Regression (GWR) function is executed with the parameters adaptive equals false and longlat equals false, the model treats the data as being in a projected planar coordinate system. Consequently, the bandwidth produced represents a fixed distance measured in metres. In this case, the recommended bandwidth of approximately 971 metres indicates that the model considers nearby data points within roughly a one-kilometre radius when estimating local spatial relationships. If the dataset were instead based on a geographic coordinate system using latitude and longitude, and the longlat parameter was set to true, the bandwidth would have been expressed in degrees.
:::

### GWModel method - fixed bandwith

Now we can use the code chunk below to calibrate the gwr model using fixed bandwidth and gaussian kernel.

```{r}
gwr.fixed <- gwr.basic(
  formula = SELLING_PRICE ~ AREA_SQM + AGE  + 
    PROX_CBD + PROX_CHILDCARE + PROX_ELDERLYCARE    +
    PROX_URA_GROWTH_AREA + PROX_MRT + PROX_PARK +
    PROX_PRIMARY_SCH + PROX_SHOPPING_MALL   + 
    PROX_BUS_STOP   + NO_Of_UNITS + FAMILY_FRIENDLY + 
    FREEHOLD, 
  data=condo_resale.sf, 
  bw=bw.fixed, 
  kernel = 'gaussian', 
  longlat = FALSE)
```

The output is saved in a list of class “gwrm”. The code below can be used to display the model output.

```{r}
gwr.fixed
```

The report shows that the AICc of the gwr is 42263.61 which is significantly smaller than the globel multiple linear regression model of 42967.1.

## **Building Adaptive Bandwidth GWR Model**

In this section, we will calibrate the gwr-based hedonic pricing model by using adaptive bandwidth approach.

### Computing the adaptive bandwidth

Similar to the earlier section, we will first use bw.gwr() to determine the recommended data point to use.

The code chunk used look very similar to the one used to compute the fixed bandwidth except the adaptive argument has changed to TRUE.

```{r}
bw.adaptive <- bw.gwr(
  formula = SELLING_PRICE ~ AREA_SQM + AGE  + 
    PROX_CBD + PROX_CHILDCARE + PROX_ELDERLYCARE    +
    PROX_URA_GROWTH_AREA + PROX_MRT + PROX_PARK +
    PROX_PRIMARY_SCH + PROX_SHOPPING_MALL   + 
    PROX_BUS_STOP   + NO_Of_UNITS + 
    FAMILY_FRIENDLY + FREEHOLD, 
  data=condo_resale.sf, 
  approach="CV", 
  kernel="gaussian", 
  adaptive=TRUE, 
  longlat=FALSE)
```

The result shows that the 30 is the recommended data points to be used.

### Constructing the adaptive bandwidth gwr model

Now, we can go ahead to calibrate the gwr-based hedonic pricing model by using adaptive bandwidth and gaussian kernel as shown in the code chunk below.

```{r}
gwr.adaptive <- gwr.basic(
  formula = SELLING_PRICE ~ AREA_SQM + AGE + 
    PROX_CBD + PROX_CHILDCARE + PROX_ELDERLYCARE +
    PROX_URA_GROWTH_AREA + PROX_MRT + PROX_PARK +
    PROX_PRIMARY_SCH + PROX_SHOPPING_MALL + PROX_BUS_STOP +
    NO_Of_UNITS + FAMILY_FRIENDLY + FREEHOLD,
  data=condo_resale.sf, 
  bw=bw.adaptive, 
  kernel = 'gaussian', 
  adaptive=TRUE, 
  longlat = FALSE)
```

The code below can be used to display the model output.

```{r}
gwr.adaptive
```

The report shows that the AICc the adaptive distance gwr is 41982.22 which is even smaller than the AICc of the fixed distance gwr of 42263.61.

## **Visualising GWR Output**

Besides the regression residuals, the output table also includes several important fields. These include the observed and predicted y values, the condition number, the local R² values, the residuals and the coefficients and standard errors for the explanatory variables.

-   **Condition Number:** This value checks for local multicollinearity, which happens when variables are too closely related. A condition number above 30 suggests the results may be unstable or unreliable.

-   **Local R²:** Shows how well the model fits the observed data in each area. Values closer to 1 mean a better fit, while low values mean the model performs poorly.

-   **Predicted Values:** These are the estimated values produced by the model based on the input variables.

-   **Residuals:** Shows the difference between the observed and predicted values. Standardized residuals make comparison easier and can be mapped to show areas of under or overprediction.

-   **Coefficient Standard Errors:** Measures how reliable each coefficient is. Smaller values mean higher confidence, while large ones may indicate local multicollinearity problems.

They are all stored in a SpatialPointsDataFrame or SpatialPolygonsDataFrame object integrated with fit.points, GWR coefficient estimates, y value, predicted values, coefficient standard errors and t-values in its “data” slot in an object called **SDF** of the output list.

## **Converting SDF into *sf* data.frame**

To visualise the fields in **SDF**, we need to first covert it into **sf** data.frame by using the code chunk below.

```{r}
condo_resale.sf.adaptive <-
  st_as_sf(gwr.adaptive$SDF) %>%
  st_transform(crs=3414)
```

Next, *glimpse()* is used to display the content of *condo_resale.sf.adaptive* sf data frame.

```{r}
glimpse(condo_resale.sf.adaptive)
```

```{r}
summary(gwr.adaptive$SDF$yhat)
```

## **Visualising local R²**

The code chunks below is used to create an interactive point symbol map.

```{r}
tmap_mode("plot")
tm_shape(mpsz_svy21)+
  tm_polygons(fill_alpha = 0.1) +
tm_shape(condo_resale.sf.adaptive) +  
  tm_dots(col = "Local_R2",
          border.col = "gray60",
          border.lwd = 1) +
  tm_view(set.zoom.limits = c(11,14))

tmap_mode("plot")
```

## **Visualising coefficient estimates**

The code chunks below is used to create an interactive point symbol map.

```{r}
tmap_mode("plot")
AREA_SQM_SE <- tm_shape(mpsz_svy21)+
  tm_polygons(alpha = 0.1) +
tm_shape(condo_resale.sf.adaptive) +  
  tm_dots(col = "AREA_SQM_SE",
          border.col = "gray60",
          border.lwd = 1) +
  tm_view(set.zoom.limits = c(11,14))

AREA_SQM_TV <- tm_shape(mpsz_svy21)+
  tm_polygons(alpha = 0.1) +
tm_shape(condo_resale.sf.adaptive) +  
  tm_dots(col = "AREA_SQM_TV",
          border.col = "gray60",
          border.lwd = 1) +
  tm_view(set.zoom.limits = c(11,14))

tmap_arrange(AREA_SQM_SE, AREA_SQM_TV, 
             asp=1, ncol=2,
             sync = TRUE)

tmap_mode("plot")
```

# **Reference**

-   [Kam, T. S. Geospatial Data Science with R. R for Geospatial Data Science and Analytics](https://r4gdsa.netlify.app/chap13.html#visualising-gwr-output)
