---
title: "In-class Ex08"
format: html
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Packages Used

```{r}
pacman::p_load(httr, tidyverse, sf, tmap, jsonlite, progress, spdep, GWmodel, SpatialML, rsample, Metrics, knitr, kableExtra, spatialRF, randomForestExplainer)
```

# Dataset

**HDB Resale data**: a list of HDB resale transacted prices in Singapore from Jan 2017 onwards. It is in csv format which can be downloaded from Data.gov.sg.

# Geocoding HDB resale data

## Reading the CSV file

*read.csv()* will be used to extract the data into an object called *HDBresale*.

```{r}
HDBresale <- read.csv("data/rawdata/HDBresale.csv") %>%
  filter(flat_type == "4 ROOM") %>%
  filter(month >= "2025-09" & month < "2025-10")
```

```{r}
#|eval: false
#|echo: false
write_rds(HDBresale, "data/rds/HDBresale.rds")
```

```{r}
#|echo: false
HDBresale <- read_rds("data/rds/HDBresale.rds")
```

```{r}
HDBresale$street_name <- gsub(
  "ST\\.", "SAINT",
  HDBresale$street_name)
```

```{r}
#|eval: false
geocode <- function(block, streetname) {
  base_url <- "https://onemap.gov.sg/api/common/elastic/search"
  address <- paste(block, streetname, sep = " ")
  query <- list(
    "searchVal" = address,
    "returnGeom" = "Y",
    "getAddrDetails" = "N",
    "pageNum" = "1"
  )

  res <- GET(base_url, query = query)
  restext <- content(res, as = "text")

  output <- fromJSON(restext) %>%
    as.data.frame() %>%
    select(results.LATITUDE, results.LONGITUDE)

  return(output)
}
```

```{r}
#|eval: false
HDBresale$LATITUDE <- 0
HDBresale$LONGITUDE <- 0

for (i in 1:nrow(HDBresale)){
  temp_output <- geocode(HDBresale[i,4],HDBresale[i,5])
  HDBresale$LATITUDE[i] <- temp_output$results.LATITUDE
  HDBresale$LONGITUDE[i] <- temp_output$results.LONGITUDE
}
```

Convert our object into sf format in SVY21. As the original crs is in 4326, we use *st_transform()* to convert into SVY21 format.

```{r}
HDBresale_sf <- HDBresale %>%
  st_as_sf(
    coords = c("LONGITUDE", "LATITUDE"),
    crs=4326) %>%
      st_transform(crs = 3414)
```

```{r}
st_crs(HDBresale_sf)
head(HDBresale_sf)
```

Extracts the preschool locations from the kml file and converts the data into SVY21 format.

```{r}
preschool = st_read("data/rawdata/PreschoolsLocation.kml"
                    )%>% st_transform(crs = 3414)
```

To calculate the number of preschools within 350m for each unit.

```{r}
within_350m <- st_is_within_distance(
  HDBresale_sf, preschool, dist = 350)

HDBresale_sf <- HDBresale_sf %>%
  mutate(prischool_350m = map_int(within_350m, length))
```

# Data import

```{r}
mdata <- read_rds("data/rawdata/mdata.rds")
```

## Data sampling

Takes out a portion of a data so that we may be able to test our codes first without the need to use all the data which requires more time.

```{r}
set.seed(1234)
HDB_sample <- mdata %>%
  sample_n(1500)
```

## Checking for overlapping data

Due to the nature of our HDB layout, we may have multiple transactions within the same block and will overlap each other. It creates a column called "overlap" and shows if there is an overlap or not.

```{r}
overlapping_points <- HDB_sample %>%
  mutate(overlap = length(st_equals(., .)) > 1)
```

Due to the overlap, while working on the model we might encounter crashes due to the overlap. Hence we can use *st_jitter()* to shift the overlapping points by 1m so as to remove the issues it may cause.

```{r}
HDB_sample <- HDB_sample %>%
  st_jitter(amount = 1)
```

## Splitting of data for our model

We will split our data into training and test data respectively.

```{r}
set.seed(1234)
resale_split <- initial_split(HDB_sample,
                              prop = 6.67/10,)
train_data <- training(resale_split)
test_data <- testing(resale_split)
```

```{r}
#| eval: false
price_mlr <- lm(resale_price ~ floor_area_sqm +
                  storey_order + remaining_lease_mths +
                  PROX_CBD + PROX_ELDERLYCARE + PROX_HAWKER +
                  PROX_MRT + PROX_PARK + PROX_MALL + 
                  PROX_SUPERMARKET + WITHIN_350M_KINDERGARTEN +
                  WITHIN_350M_CHILDCARE + WITHIN_350M_BUS +
                  WITHIN_1KM_PRISCH,
                data=train_data)
olsrr::ols_regress(price_mlr)
```

```{r}
#| eval: false
#| include: false
bw_adaptive <- bw.gwr(resale_price ~ floor_area_sqm +
                  storey_order + remaining_lease_mths +
                  PROX_CBD + PROX_ELDERLYCARE + PROX_HAWKER +
                  PROX_MRT + PROX_PARK + PROX_MALL + 
                  PROX_SUPERMARKET + WITHIN_350M_KINDERGARTEN +
                  WITHIN_350M_CHILDCARE + WITHIN_350M_BUS +
                  WITHIN_1KM_PRISCH,
                  data=train_data,
                  approach="CV",
                  kernel="gaussian",
                  adaptive=TRUE,
                  longlat=FALSE)
```

```{r}
#| eval: false
#| include: false
write_rds(bw_adaptive, "data/rds/bw_adaptive.rds")
```

```{r}
#| echo: false
bw_adaptive <- read_rds("data/rds/bw_adaptive.rds")
```

```{r}
bw_adaptive
```

```{r}
gwr_adaptive <-read_rds("data/rds/bw_adaptive.rds")
```
