---
title: "In-class_Ex05: Global and Local Measures of Spatial Autocorrelation"
format: html
---

# **Installing and Loading the R packages**

::: panel-tabset
## Packages Used

-   **sf** Provides the core tools for handling spatial data
-   **tmap** A package used for producing maps for visualisation
-   **tidyverse** A collection of R packages designed for data science tasks such as data import, cleaning, transformation, and visualisation
-   **spdep** Provides functions for spatial dependence and spatial regression analysis, including tools for creating spatial weights, testing for spatial autocorrelation, and fitting spatial econometric models

## Code Chunk

The code chunk below uses *p_load()* of pacman package to check if the packages are installed in the computer. If they are, then they will be launched into R.

```{r}
pacman::p_load(sf, sfdep, tmap, tidyverse)
```
:::

# The Data

For this purpose of this in-class exercise, the Hunan datasets will be used.

::: panel-tabset
## **Importing hunan files**

The following code chunk will be using the *st_read()* function of sf package to import the Hunan shapefile into our R object, **hunan**.

```{r}
hunan <- st_read(dsn = "data/geospatial", 
                 layer = "Hunan")
```

## Importing csv files

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```
:::

## **Performing relational join**

After the importing of the files into the R environment, we will proceed to join the files.

The code chunk below will be used to update the attribute table of *hunan*’s SpatialPolygonsDataFrame with the attribute fields of *hunan2012* dataframe. This is performed by using *left_join()* of **dplyr** package.

```{r}
hunan_GDPPC <- left_join(hunan,hunan2012) %>%
  select(1:4, 7, 15)
```

# **Plotting a choropleth map**

## Deriving Queen's contiguity weights: sfdep methods

-   nb: A neighbour list object as created by `st_neighbours()`.

-   style: Default "W" for row standardized weights. This value can also be "B", "C", "U", "minmax" and "S". B is the basic binary coding, W is row standardised (sums over all links to n), C is globally standardised (sums over all links to n), U is equal to C divided by the number of neighbours (sums over all links to unity), while S is the variance-stabilizing coding scheme proposed by Tiefelsdorf et al. 1999, p. 167-168 (sums over all links to n).

-   allow_zero: If TRUE, assigns zero as lagged value to zone without neighbors.

```{r}
wm_q <- hunan_GDPPC %>%
  mutate(nb = st_contiguity(geometry),
         wt = st_weights(nb,
                         style = "W"),
         .before = 1)
```

## **Performing Global Moran's I Permutation Test**

In practice, Monte Carlo simulation should be used to perform the statistical test. For **sfdep**, it is supported by `global_moran_perm()`

```{r}
set.seed(1234)
```

Next, `global_moran_perm()` is used to perform Monte Carlo simulation.

```{r}
global_moran_perm(wm_q$GDPPC,
                  wm_q$nb,
                  wm_q$wt,
                  nsim = 99)
```

The statistical report on the results shows that the p-value is smaller than alpha value of 0.05. Hence we have enough statistical evidence to reject the null hypothesis that the spatial distribution of GDP per captia resemble random distribution (i.e.

## LISA Map
