---
title: "In-class Ex09 - Modelling Geographical Accessibility"
format: html
execute:
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Introduction

This exercise shows how to model geographical accessibility in R using standard geospatial packages.\
We will load, clean, and visualise spatial data, and calculate accessibility using several methods.

# The Data

You’ll work with four datasets:

-   MP14_SUBZONE_NO_SEA_PL: URA Master Plan 2014 subzones (data.gov.sg)\
-   hexagons: 250 m radius grid created with sf::st_make_grid()\
-   ELDERCARE: point layer of eldercare facilities (data.gov.sg)\
-   OD_Matrix: CSV file with distance and cost values in metres:
    -   origin_id and destination_id
    -   entry_cost, network_cost, exit_cost
    -   total_cost = entry_cost + network_cost + exit_cost

# **Installing and Loading the R packages**

::: panel-tabset
## Packages Used

The R packages need for this exercise are as follows:

-   Spatial data handling

    -   sf

-   Modelling geographical accessibility

    -   spatialAcc

-   Attribute data handling

    -   tidyverse, especially readr and dplyr

-   thematic mapping

    -   tmap

-   Staistical graphic

    -   ggplot2

-   Statistical analysis

    -   ggstatsplot

## Code Chunk

The code chunk below uses *p_load()* of pacman package to check if the packages are installed in the computer. If they are, then they will be launched into R.

```{r}
pacman::p_load(tmap, SpatialAcc, sf, 
               ggstatsplot, reshape2,
               tidyverse)
```
:::

# Geospatial Data Wrangling

This section focuses on getting our spatial data ready for analysis. We begin by importing the layers we need, such as the URA subzones, the hexagonal grid that defines our study area, and the eldercare facility locations. These files come in shapefile format and will later be combined to analyse accessibility across Singapore.

## Importing Geospatial Data

We start by reading the shapefiles using the **sf** package in R. Loading them as spatial objects allows us to use them in mapping and analysis. It is also a good practice to check that each file has been imported correctly and that its geometry type is properly recognised.

```{r}
mpsz <- st_read(dsn = "data/geospatial", layer = "MP14_SUBZONE_NO_SEA_PL")
hexagons <- st_read(dsn = "data/geospatial", layer = "hexagons") 
eldercare <- st_read(dsn = "data/geospatial", layer = "ELDERCARE") 
```

## Updating CRS Information

To make sure every dataset aligns correctly on the map, we convert them all to the same coordinate reference system, which is SVY21 (EPSG:3414). This ensures that all coordinates are using the same reference and can be compared accurately. After converting, we confirm that the transformation has worked as expected.

```{r}
mpsz <- st_transform(mpsz, 3414)
eldercare <- st_transform(eldercare, 3414)
hexagons <- st_transform(hexagons, 3414)
st_crs(mpsz)
```

## Cleaning and Updating Attribute Fields

Next, we clean up the datasets by keeping only the necessary columns. We also create two additional fields to support our analysis. The first is capacity, which shows how much service an eldercare centre can provide. The second is demand, which reflects the level of need in each hexagon.\

For this exercise, both values are set to 100 as placeholders. In a real-world project, they would be based on actual data such as population or service capacity.

```{r}
eldercare <- eldercare %>%
  select(fid, ADDRESSPOS) %>%
  mutate(capacity = 100)

hexagons <- hexagons %>%
  select(fid) %>%
  mutate(demand = 100)
```

# Aspatial Data Handling and Wrangling

Here, we prepare the non-spatial dataset known as the Origin-Destination (OD) matrix. This dataset records the travel distances between every demand location and every eldercare facility. It is an essential input for accessibility modelling since it tells us how far people must travel to reach each service.

## Importing Distance Matrix

We load the OD matrix from a CSV file. Each record represents a pair of points: an origin (a hexagon) and a destination (an eldercare centre). The matrix includes the total travel distance between them, measured in metres. This gives us a full picture of how connected different parts of the study area are to eldercare services.

```{r}
ODMatrix <- read_csv("data/aspatial/OD_Matrix.csv", skip = 0)

ODMatrix
```

## Tidying Distance Matrix

The imported data starts in a long format, meaning that each row shows one origin and one destination. For modelling, we need it in a wide format, where each row represents one origin and each column represents one destination. The value inside each cell shows the distance between them.

```{r}
distmat <- ODMatrix %>%
  select(origin_id, destination_id, total_cost) %>%
  spread(destination_id, total_cost)%>%
  select(c(-c('origin_id')))
```

After restructuring, we convert the distances from metres to kilometres to make the numbers easier to interpret.

```{r}
distmat_km <- as.matrix(distmat/1000)
```

# Modelling and Visualising Accessibility using Hansen Method

The Hansen method is a classic way to measure accessibility. It looks at both the number of available services and their distance from the population. Areas that are close to many facilities will have higher accessibility scores, while areas that are far away will have lower scores.

## Computing Hansen’s Accessibility

We calculate accessibility using the **SpatialAcc** package. The function takes into account the demand, the capacity of eldercare centres, and the distance matrix. It produces an accessibility score for each hexagon. These scores are then joined back to the spatial grid so that they can be visualised on a map.

```{r}
acc_Hansen <- data.frame(ac(hexagons$demand,
                            eldercare$capacity,
                            distmat_km, 
                            #d0 = 50,
                            power = 2, 
                            family = "Hansen"))
colnames(acc_Hansen) <- "accHansen"
acc_Hansen <- as_tibble(acc_Hansen)
hexagon_Hansen <- bind_cols(hexagons, acc_Hansen)
```

The default field name is very messy, we will rename it to accHansen by using the code chunk below.

```{r}
colnames(acc_Hansen) <- "accHansen"
acc_Hansen
```

Next, we will convert the data table into tibble format by using the code chunk below.

```{r}
acc_Hansen <- as_tibble(acc_Hansen)

```

Lastly, *bind_cols()* of **dplyr** will be used to join the acc_Hansen tibble data frame with the hexagons simple feature data frame. The output is called *hexagon_Hansen*.

```{r}
hexagon_Hansen <- bind_cols(hexagons, acc_Hansen)
hexagon_Hansen
```

Actually, the steps above can be perform by using a single code chunk as shown below.

```{r}
#| eval: false
acc_Hansen <- data.frame(ac(hexagons$demand,
                            eldercare$capacity,
                            distmat_km, 
                            #d0 = 50,
                            power = 0.5, 
                            family = "Hansen"))

colnames(acc_Hansen) <- "accHansen"
acc_Hansen <- as_tibble(acc_Hansen)
hexagon_Hansen <- bind_cols(hexagons, acc_Hansen)
```

::: panel-tabset
## power 0.5

```{r}
acc_Hansen_05 <- data.frame(ac(hexagons$demand,
                               eldercare$capacity,
                               distmat_km, 
                               power = 0.5, 
                               family = "Hansen"))
colnames(acc_Hansen_05) <- "accHansen_05"
acc_Hansen_05 <- as_tibble(acc_Hansen_05)
hexagon_Hansen_05 <- bind_cols(hexagons, acc_Hansen_05)
```

```{r}
mapex <- st_bbox(hexagons)
tmap_mode("plot")

plot1 <- tm_shape(hexagon_Hansen_05, bbox = mapex) + 
  tm_polygons(fill = "accHansen_05",
              fill.scale = tm_scale_intervals(style="quantile", n=10, values="brewer.blues")) +
  tm_shape(eldercare) + tm_dots(size=0.3, fill="red", col="black") +
  tm_title("Power = 0.5") + tm_layout(frame=TRUE) +
  tm_compass(type="8star", size=2) + tm_scalebar(position=c("right","bottom")) +
  tm_grid(alpha=0.2)
plot1
```

## power 1.0

```{r}
acc_Hansen_1 <- data.frame(ac(hexagons$demand,
                              eldercare$capacity,
                              distmat_km, 
                              power = 1, 
                              family = "Hansen"))
colnames(acc_Hansen_1) <- "accHansen_1"
acc_Hansen_1 <- as_tibble(acc_Hansen_1)
hexagon_Hansen_1 <- bind_cols(hexagons, acc_Hansen_1)
```

```{r}
plot2 <- tm_shape(hexagon_Hansen_1, bbox = mapex) + 
  tm_polygons(fill = "accHansen_1",
              fill.scale = tm_scale_intervals(style="quantile", n=10, values="brewer.blues")) +
  tm_shape(eldercare) + tm_dots(size=0.3, fill="red", col="black") +
  tm_title("Power = 1.0") + tm_layout(frame=TRUE) +
  tm_compass(type="8star", size=2) + tm_scalebar(position=c("right","bottom")) +
  tm_grid(alpha=0.2)
plot2
```

## power 1.5

```{r}
acc_Hansen_15 <- data.frame(ac(hexagons$demand,
                                eldercare$capacity,
                                distmat_km, 
                                power = 1.5, 
                                family = "Hansen"))
colnames(acc_Hansen_15) <- "accHansen_15"
acc_Hansen_15 <- as_tibble(acc_Hansen_15)
hexagon_Hansen_15 <- bind_cols(hexagons, acc_Hansen_15)
```

```{r}
plot3 <- tm_shape(hexagon_Hansen_15, bbox = mapex) + 
  tm_polygons(fill = "accHansen_15",
              fill.scale = tm_scale_intervals(style="quantile", n=10, values="brewer.blues")) +
  tm_shape(eldercare) + tm_dots(size=0.3, fill="red", col="black") +
  tm_title("Power = 1.5") + tm_layout(frame=TRUE) +
  tm_compass(type="8star", size=2) + tm_scalebar(position=c("right","bottom")) +
  tm_grid(alpha=0.2)
plot3
```

## power 2.0

```{r}
acc_Hansen_2 <- data.frame(ac(hexagons$demand,
                              eldercare$capacity,
                              distmat_km, 
                              power = 2, 
                              family = "Hansen"))
colnames(acc_Hansen_2) <- "accHansen_2"
acc_Hansen_2 <- as_tibble(acc_Hansen_2)
hexagon_Hansen_2 <- bind_cols(hexagons, acc_Hansen_2)
```

```{r}
plot4 <- tm_shape(hexagon_Hansen_2, bbox = mapex) + 
  tm_polygons(fill = "accHansen_2",
              fill.scale = tm_scale_intervals(style="quantile", n=10, values="brewer.blues")) +
  tm_shape(eldercare) + tm_dots(size=0.3, fill="red", col="black") +
  tm_title("Power = 2.0") + tm_layout(frame=TRUE) +
  tm_compass(type="8star", size=2) + tm_scalebar(position=c("right","bottom")) +
  tm_grid(alpha=0.2)
plot4
```

## power 2.5

```{r}
acc_Hansen_25 <- data.frame(ac(hexagons$demand,
                                eldercare$capacity,
                                distmat_km, 
                                power = 2.5, 
                                family = "Hansen"))
colnames(acc_Hansen_25) <- "accHansen_25"
acc_Hansen_25 <- as_tibble(acc_Hansen_25)
hexagon_Hansen_25 <- bind_cols(hexagons, acc_Hansen_25)
```

```{r}
plot5 <- tm_shape(hexagon_Hansen_25, bbox = mapex) + 
  tm_polygons(fill = "accHansen_25",
              fill.scale = tm_scale_intervals(style="quantile", n=10, values="brewer.blues")) +
  tm_shape(eldercare) + tm_dots(size=0.3, fill="red", col="black") +
  tm_title("Power = 2.5") + tm_layout(frame=TRUE) +
  tm_compass(type="8star", size=2) + tm_scalebar(position=c("right","bottom")) +
  tm_grid(alpha=0.2)
  
plot5
```
:::
